Solving the Poisson-equation in one dimension \\ \hrulefill\small{ FYS3150: Computational Physics }\hrulefill

### 2016-09-14 Project1/article/article.tex
Scrutiny of program. Algorithm part in article almost done.

\usepackage[]{graphicx}
\usepackage[]{hyperref}
\usepackage[]{physics}

\usepackage[T1]{fontenc}

@@

By defining $\vb{f}=h^2f_i$ one can rewrite equation
\ref{eq:2approx} as $-v_{i+1}-2v_i+v_{i-1}=h^2f_i$. If we ignore the
end points, $i=0$ and $i=n+1$, this equation can be represented as a
matrix equation.
\begin{equation}
A\vb{v}=\vb{f}
\end{equation}
\begin{equation}
\begin{bmatrix}
2 & -1 & 0 & 0 & \cdots & 0 & 0 & 0 \\
-1 & 2 & -1 & 0 & \cdots & 0 & 0 & 0 \\
0 & -1 & 2 & -1 & \cdots & 0 & 0 & 0 \\ 
& & \vdots &  & \ddots &  & \vdots & \\
0 & 0 & 0 & 0 & \cdots & -1 & 2 & -1 \\
0 & 0 & 0 & 0 & \cdots & 0 & -1 & 2 
\end{bmatrix}
\begin{bmatrix}
v_1 \\
v_2 \\
v_3 \\
\vdots \\
v_{n-1} \\
v_n
\end{bmatrix}=
\begin{bmatrix}
f_1 \\
f_2 \\
f_3 \\
\vdots \\
f_{n-1} \\
f_n
\end{bmatrix}
\end{equation}

@@

\subsection{Tridiagonal Matrix Algorithm}
Our tridiagonal system can be represented by 
\begin{equation}
a_ix_{i-1} + b_ix_i + c_ix_{i+1} = \vb{f}
\end{equation}
with $a_i=-1,b_i=2$ and $c_i=-1$, except for $a_1=0$ and $c_n=0$. Row
reducing a matrix will reveal how the algorithm functions. Limiting
the problem to four dimensions for easier reading and to save the
rainforest\footnote{Writing out a general case will take up more paper
space}.

@@

This pattern can be summarized nicely by the following
difference equations. This computation will from now on be referred to
as the ``forward substitution''. 
\begin{equation}
\beta_i=b_i-\frac{c_{i-1}}{\beta_{i-1}}, \quad
\tilde{f}_i=(\beta_i-\tilde{f}_{i-1}a_i)/\beta_i, \quad i \in [2,n], \quad
\end{equation}

@@

The notation is updated again to $v_3 =\tilde{f}_3 -
\frac{c_3}{\beta_3}\tilde{f}_4$. By continuing this iterative process
all the was to the beginning of the matrix by backward substitution
leaves us with a new recurence relation.
\begin{equation}
v_i=\tilde{f}_i - \frac{c_i}{\beta_i}\tilde{f}_{i+1}, \quad i \in \{n-1,n-2,\dots,1\}
\end{equation}

@@

>>>>

\usepackage[]{graphicx}
\usepackage[]{hyperref}
\usepackage[]{physics}
\usepackage[]{listings}
\usepackage[T1]{fontenc}
\usepackage{color}

\definecolor{mygreen}{rgb}{0,0.6,0}
\definecolor{mygray}{rgb}{0.5,0.5,0.5}
\definecolor{mymauve}{rgb}{0.58,0,0.82}

\lstset{ %
  backgroundcolor=\color{white},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}
  basicstyle=\footnotesize,        % the size of the fonts that are used for the code
  breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
  breaklines=true,                 % sets automatic line breaking
  captionpos=b,                    % sets the caption-position to bottom
  commentstyle=\color{mygreen},    % comment style
  deletekeywords={...},            % if you want to delete keywords from the given language
  escapeinside={\%*}{*)},          % if you want to add LaTeX within your code
  extendedchars=true,              % lets you use non-ASCII characters; for 8-bits encodings only, does not work with UTF-8
  frame=single,	                   % adds a frame around the code
  keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
  keywordstyle=\color{blue},       % keyword style
  language=c++,                 % the language of the code
  otherkeywords={*,...},           % if you want to add more keywords to the set
  rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
  showspaces=false,                % show spaces everywhere adding particular underscores; it overrides 'showstringspaces'
  showstringspaces=false,          % underline spaces within strings only
  showtabs=false,                  % show tabs within strings adding particular underscores
  stepnumber=2,                    % the step between two line-numbers. If it's 1, each line will be numbered
  stringstyle=\color{mymauve},     % string literal style
  tabsize=2,	                   % sets default tabsize to 2 spaces
}

@@

By defining $\vb{f}=h^2f_i$ one can rewrite equation
\ref{eq:2approx} as $-v_{i+1}-2v_i+v_{i-1}=h^2f_i$. If we ignore the
end points, $i=0$ and $i=n+1$, this equation can be represented as a
matrix equation.
\begin{equation}
\label{eq:matrix}
A\vb{v}=\vb{f}
\end{equation}
\begin{equation}
\begin{bmatrix}
2 & -1 & 0 & 0 & \cdots & 0 & 0 & 0 \\
-1 & 2 & -1 & 0 & \cdots & 0 & 0 & 0 \\
0 & -1 & 2 & -1 & \cdots & 0 & 0 & 0 \\ 
& & \vdots &  & \ddots &  & \vdots & \\
0 & 0 & 0 & 0 & \cdots & -1 & 2 & -1 \\
0 & 0 & 0 & 0 & \cdots & 0 & -1 & 2 
\end{bmatrix}
\begin{bmatrix}
v_1 \\
v_2 \\
v_3 \\
\vdots \\
v_{n-1} \\
v_n
\end{bmatrix}=
\begin{bmatrix}
f_1 \\
f_2 \\
f_3 \\
\vdots \\
f_{n-1} \\
f_n
\end{bmatrix}
\end{equation}

@@

\subsection{Tridiagonal Matrix Algorithm}
Our tridiagonal system can be represented by 
\begin{equation}
a_ix_{i-1} + b_ix_i + c_ix_{i+1} = \vb{f}
\end{equation}
with $a_i=-1,b_i=2$ and $c_i=-1$, except for $a_1=0$ and $c_n=0$. Row
reducing a matrix will reveal how the algorithm functions. Limiting
the problem to four dimensions for easier reading and to save the
rainforest\footnote{Writing out a general case will also take up more paper
space}.

@@

This pattern can be summarized nicely by the following
difference equations. This computation will from now on be referred to
as the ``forward substitution''. 
\begin{equation}
\label{eq:forward1}
\beta_i=b_i-\frac{a_ic_{i-1}}{\beta_{i-1}}, \quad
\tilde{f}_i=f_i-a_i\tilde{f}_{i-1}/\beta_{i-1}, \quad i \in [2,n], \quad
\end{equation}

In c++, this part of the algorithm is implemented just below. Notice
that $b_i$ is overwritten instead of initializing a seperate vector
for $\beta$. 
\lstinputlisting[language=c++, firstline=65, lastline=71]{../problems.cpp}

@@

The notation is updated again to $v_3 =\tilde{f}_3 -
\frac{c_3}{\beta_3}\tilde{f}_4$. By continuing this iterative process
all the was to the beginning of the matrix by backward substitution
leaves us with a second recurrence relation.
\begin{equation}
\label{eq:backward1}
v_i=(\tilde{f}_i - c_iv_{i+1})/\beta_i, \quad i \in \{n-1,n-2,\dots,1\}
\end{equation}

The backward substitution is implemented in C++ like so:
\lstinputlisting[language=c++, firstline=73,
lastline=77]{../problems.cpp}

An ever-important aspect of any algorithm is how efficient the algorithm is. It
is therefor worthwhile to consider how many floating point operations
the algorithm requires\footnote{The number of $+$,$-$,$*$ and $/$
  operations are counted}. For the forward substitution there are $6$
FLOPS in each iteration and $n-1$ iterations, resulting in $6(n-1)$
FLOPS for the entire loop. Additionally, $1(n-1)$ FLOPS are required
for the backward substitution, adding up to 
\begin{equation}
N_{tridiag}=8(n-1)=\mathcal{O} (8n).
\end{equation}

\subsubsection{Optimizing the algorithm}

Our case is special, because the elements in each of the diagonal arrays in the
tridiagonal matrix, $A$, are equal. $a_i=-1,b_i=2$ and $c_i=-1$,
except for $a_1=0$ and $c_n=0$. This is an advantage, because the
algorithm can be adjusted, and the number of FLOPS consequently reduced.

The algorithm for the forward substitution, in equation
\ref{eq:forward1}, can be simplified to
\begin{equation}
\beta_i= b_i - 1/\beta_{i-1} \quad \tilde{f}_i=f_i-\tilde{f}_{i-1}/\beta_{i-1}
\end{equation}
and the backward substitution, in equation \ref{eq:backward1}, can be
simplified to
\begin{equation}
v_i = (\tilde{f}_i+v_{i+1})/\beta_i
\end{equation}

This results in some new and sleeker C++ code
\lstinputlisting[language=c++, firstline=146,
lastline=158]{../problems.cpp}

One can see that this new and improved algorithm only requires
$6(n-1)$ FLOPS. In general, a Gaussian row reduction algorithm
requires $\frac{2}{3}n^3+\mathcal{O}(n^2)$ FLOPS \cite{morten}, a
great improvement!

@@

The LU decompositon method is a method where the matrix $A$ in
equation \ref{eq:matrix} ($A\vb{v}=\vb{f}$) can be rewritten as a product of two matrices
$L$ and $U$, where $L$ is lower triangular and and $U$ is upper
triangular. This method only works for non-singular, invertible
matrices, which the matrix $A$ certainly is.
